Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-05-19 manuscript.tex
Figure changes and acknowledgements

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=1.0\textwidth]{Fig2.pdf}
	\end{center}
\end{figure}
\begin{figure}[ht!]
        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs for AT-all, GC-end, GC-core, and GC-mix. (a) \textbf{Macrostates.} Schematic representation of the seven metastable macrostates occupied by one or more of the four sequences: fully hybridized state (H), two-base out-of-register shifted states in the 5$^\prime$ (5S2) or 3$^\prime$ (3S2) direction, four-base out-of-register shifted states in the 5$^\prime$ (5S4) or 3$^\prime$ (3S4) direction, and the fully dissociated state (D). The line drawings represent the 10-base self-complementary sequences, where red-to-blue contacts indicates (possible) WC base pairing and black indicates an unbound bases. Adjacent to each line drawing we provide representative molecular structures corresponding to that macrostate. (b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities. Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts. \noter{Add percentages to blue bars in white text.} (c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence. The orange circles correspond to the macrostates occupied by each sequence and are labeled by the macrostate codes reported in panel a. The radii of the circles is proportional to the logarithm of the equilibrium occupancy probabilities reported in panel b. Molecular renderings of an illustrative snapshot from the coarse-grained molecular simulations are provided next to each macrostate. The grey arrows between macrostates indicate the presence of a probability flux between this pair of states at equilibrium and the arrow thickness is proportional to the flux. (To avoid congesting the diagram, arrows are not reported for probability fluxes lower than $3\times10^{-6}$.) The numerical value overlaid on each arrow reports the conditional probability that a system occupying the macrostate at the start of the arrow at time $t$ will transition to the macrostate at the end of the arrow by time $(t+\tau)$, where $\tau$=1.2 ns is the lag time corresponding to a single time step of the MSM. Large orange circles correspond to thermodynamically favorable states and large grey arrows correspond to kinetically favorable transitions. (d) \textbf{Kinetic time scales.} Distribution of MSM implied time scales for each sequence. The leading implied time scale corresponds to the characteristic time scale for hybridization/dehybridization and is approximately equal for all systems since simulations were conducted at the same concentration and at the respective melting temperatures. The higher order implied time scales correspond to a spectrum of kinetic relaxations between the constituent macrostates in the MSM corresponding to shifted and/or frayed states.
    }
    \label{fig:allseq_table}
\end{figure}

@@

We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of predictive sequence-dependent models for the thermodynamics of DNA association. The nearest neighbor (NN) model is popular model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends. We apply the NN model to estimate the (relative) free energy $F$ of each of the macrostates occupied by each of the four sequences. The free energy of each macrostate is related to its equilibrium occupancy probability $P$ via the statistical mechanical relationship $F = -k_B T \ln P + C$, where $T$ is temperature, $k_B$ is Boltzmann's constant, and $C$ is an additive constant reflecting our ignorance of the absolute scale of free energies. Since the additive constant cancels out in the evaluation of relative free energies between macrostates, this calculation provides a means to compare free energy differences predicted by our MSMs (Fig.~\ref{fig:allseq_table}) to those of the NN model. For clarity of presentation, we specify the additive constant $C$ in both the MSM and NN models by arbitrarily assigning the free energy of the hybridized state H to define the zero of free energy. A full accounting of the application of the NN model to the macrostates occupied by each of the four sequences is provided in the \blauw{Supporting Information}.

As illustrated in Fig.~\ref{fig:NN_table}, we see that the NN model tends to predict lower free energies for each macrostate compared to the MSM and therefore overestimate macrostate stability relative to the hybridized state by as much as 14.7 kJ/mol. The only exception to this trend is the frayed state (F4) unique to CG-core, where the MSM and NN predictions are in good agreement. Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the 5S2, 3S2, 5S4, and 3S4 states? First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data. There are approximations inherent in the 3SPN.2 model, not least of which is the coarse-grained representation that integrates over atomic degrees of freedom, and in the NN model that was fitted to limited experimental data assuming a low-order expansion in terms of nearest neighbor additive contributions. Second, although 3SPN.2 is expected to capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the model was not parameterized to fully capture the enthalpic contributions of the interaction of unbound bases with terminal base pairs, whereas this term is explicitly included within the NN model. Third, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end tend to destabilize the duplex.\citep{DiMichele2014EffectHybridization}

@@

In sum, the qualitative trends in the macrostate thermodynamic stabilities are in good agreement between the MSM and NN models, but show quantitative discrepancies for macrostates possessing inert tails. In these instances the NN model overestimates macrostate stability relative to the hybridized state by 4.0-14.7 kJ/mol. The MSM predictions are also consistent with the experimental expectation that 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the NN predictions can be in conflict with this trend.

@@

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=160mm, scale=1]{Fig3.pdf}
        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state that is assigned a free energy of zero. (b) Discrepancy between the macrostate free energy predictions of the MSM relative to the NN model. With the exception of the F4 frayed state, the NN model predicts lower free energies than the MSM predictions. \noter{(i) y-axis label in a: ``free energy relative to hybridized state, F (kJ/mol)'' (ii) y-axis label in b: ``free energy discrepancy between MSM and NN model relative to hybridized state, $\Delta$F (kJ/mol)'' -- might have to go on two lines}}
        \label{fig:NN_table}
	\end{center}
\end{figure}

@@

The GC-core MSM represents a departure from the relatively rich and complex kinetic networks dominated by out-of-register shifted states to a much simpler one dominated by fraying (Fig.~\ref{fig:allseq_table}c). The MSM contains only three states -- hybridized H, dehybridized D, and frayed F4. The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs on one side or other of the core, while the other run of four A:T pairs remains free. (As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) \noter{Maybe add a sentence here on other frayed states containing 3, 2, 1 frayed bases but not metastable within our MSM.} Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4. Previous studies would suggest that hybridization of this sequence should proceed via a zippering mechanism, wherein upon formation of the strong central G:C WC base pairings the duplex helix should rapidly assemble in a middle-out fashion.~\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results are partially consistent with this expectation, but reveal the frayed state F4 to be unexpectedly metastable, serving as a long-lived state with an mean life time of 1.8 ns. The stability of the state is attributable to the enthalpic stabilization offered by the up to six WC pairs and the entropic stabilization associated with the configurational entropy of the two free AT-tails.

@@

%We present in Fig.~\ref{fig:T-jump_stretching}b-d our T-jump IR t-HDVE difference spectra and corresponding normalized time traces at 1600 cm$^{-1}$ and 1660 cm$^{-1}$. The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations. Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$. The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature. The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ corresponds to \noter{What physical process.}  A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix. This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6. This result provides an experimental validation of the kinetic predictions of the MSM models.

@@

This material is based upon work supported by the National Science Foundation under Grant No.\ CHE-1841805. B.A. acknowledges support from the NSF Graduate Research Fellowship Program. This work was completed in part with resources provided by the University of Chicago Research Computing Center. We gratefully acknowledge computing time on the University of Chicago high-performance GPU-based cyberinfrastructure supported by the National Science Foundation under Grant No.\ DMR-1828629.

>>>>

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=1.0\textwidth]{Fig2.pdf}
	\end{center}
\end{figure}
\begin{figure}[ht!]
        \caption{Thermodynamic and kinetic predictions of the sequence-specific MSMs for AT-all, GC-end, GC-core, and GC-mix. (a) \textbf{Macrostates.} Schematic representation of the seven metastable macrostates occupied by one or more of the four sequences: fully hybridized state (H), two-base out-of-register shifted states in the 5$^\prime$ (5S2) or 3$^\prime$ (3S2) direction, four-base out-of-register shifted states in the 5$^\prime$ (5S4) or 3$^\prime$ (3S4) direction, and the fully dissociated state (D). The line drawings represent the 10-base self-complementary sequences, where red-to-blue contacts indicates (possible) WC base pairing and black indicates an unbound bases. Adjacent to each line drawing we provide representative molecular structures corresponding to that macrostate. (b) \textbf{Thermodynamics.} Histograms reporting the number of the $10^7$ total frames within the 1 ms of simulation trajectories observed to occupy each of the seven macrostates, corresponding to our numerical estimates of the equilibrium occupancy probabilities. Uncertainties are calculated across 100 MSMs using a Bayesian MSM estimation are reported for each bar and are very small compared to the total counts. (c) \textbf{Kinetic networks.} MSMs illustrating the kinetic network for each sequence. The orange circles correspond to the macrostates occupied by each sequence and are labeled by the macrostate codes reported in panel a. The radii of the circles is proportional to the logarithm of the equilibrium occupancy probabilities reported in panel b. Molecular renderings of an illustrative snapshot from the coarse-grained molecular simulations are provided next to each macrostate. The grey arrows between macrostates indicate the presence of a probability flux between this pair of states at equilibrium and the arrow thickness is proportional to the flux. (To avoid congesting the diagram, arrows are not reported for probability fluxes lower than $3\times10^{-6}$.) The numerical value overlaid on each arrow reports the conditional probability that a system occupying the macrostate at the start of the arrow at time $t$ will transition to the macrostate at the end of the arrow by time $(t+\tau)$, where $\tau$=1.2 ns is the lag time corresponding to a single time step of the MSM. Large orange circles correspond to thermodynamically favorable states and large grey arrows correspond to kinetically favorable transitions. (d) \textbf{Kinetic time scales.} Distribution of MSM implied time scales for each sequence. The leading implied time scale corresponds to the characteristic time scale for hybridization/dehybridization and is approximately equal for all systems since simulations were conducted at the same concentration and at the respective melting temperatures. The higher order implied time scales correspond to a spectrum of kinetic relaxations between the constituent macrostates in the MSM corresponding to shifted and/or frayed states.
    }
    \label{fig:allseq_table}
\end{figure}

@@

We first compare the thermodynamic predictions for the equilibrium macrostate probabilities to those of predictive sequence-dependent models for the thermodynamics of DNA association. The nearest neighbor (NN) model is a popular empirical model of DNA hybridization thermodynamics that predicts the free energy of duplex formation as a sum over helix initiation terms and the hybridization free energies of nearest neighbor pairs of bases that account for both the specific WC pairings and the modulating effects of the local (i.e., nearest neighbor) environment.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} The parameters of the NN model were estimated by regressing over 108 experimental measurements to furnish a predictive model for the free energy of association as a function of DNA sequence and explicitly account for stacking contributions of native pairs, internal mismatches, and dangling ends. We apply the NN model to estimate the (relative) free energy $F$ of each of the macrostates occupied by each of the four sequences. The free energy of each macrostate is related to its equilibrium occupancy probability $P$ via the statistical mechanical relationship $F = -k_B T \ln P + C$, where $T$ is temperature, $k_B$ is Boltzmann's constant, and $C$ is an additive constant reflecting our ignorance of the absolute scale of free energies. Since the additive constant cancels out in the evaluation of relative free energies between macrostates, this calculation provides a means to compare free energy differences predicted by our MSMs (Fig.~\ref{fig:allseq_table}) to those of the NN model. For clarity of presentation, we specify the additive constant $C$ in both the MSM and NN models by arbitrarily assigning the free energy of the hybridized state H to define the zero of free energy. A full accounting of the application of the NN model to the macrostates occupied by each of the four sequences is provided in the \blauw{Supporting Information}.

As illustrated in Fig.~\ref{fig:NN_table}, we see that the NN model tends to predict lower free energies for each macrostate compared to the MSM \noter{and therefore predicts a higher metastable state populations.} The only exception to this trend is the frayed state (F4) unique to CG-core, where the MSM and NN predictions are in good agreement. Although the trends are in qualitative agreement, what is the root of the quantitative discrepancy of the MSM and NN models in the predicted relative stability of the 5S2, 3S2, 5S4, and 3S4 states? First, the MSM is constructed bottom-up from molecularly detailed 3SPN.2 simulations whereas the NN model is fitted top-down by regression against experimental data. There are approximations inherent in the 3SPN.2 model, not least of which is the coarse-grained representation that integrates over atomic degrees of freedom, and in the NN model that was fitted to limited experimental data assuming a low-order expansion in terms of nearest neighbor additive contributions. Second, although 3SPN.2 is expected to capture some dangling end stabilization effects through base stacking and cross-stacking interactions, the model was not parameterized to fully capture the enthalpic contributions of the interaction of unbound bases with terminal base pairs, whereas this term is explicitly included within the NN model. Third, a well known deficiency of the NN model is the absence of any treatment of inert tails -- free bases that extend beyond the dangling end tend to destabilize the duplex.\citep{DiMichele2014EffectHybridization}

@@

In sum, the qualitative trends in the macrostate thermodynamic stabilities are in good agreement between the MSM and NN models, but show quantitative discrepancies for macrostates possessing inert tails. In these instances the NN model overestimates \noter{another mention of ``overestimate'' but perhaps we have qualified it more at this point in the text?} macrostate stability relative to the hybridized state by 4.0-14.7 kJ/mol. The MSM predictions are also consistent with the experimental expectation that 3$^\prime$ inert tails should be more destabilizing than the 5$^\prime$ tails, whereas the NN predictions can be in conflict with this trend.

@@

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=160mm, scale=1]{Fig3.pdf}
        \caption{Comparison of the macrostate free energy predictions of the MSMs and nearest neighbor (NN) thermodynamic model.~\citep{SantaLucia1998AThermodynamics, Santalucia2004TheMotifs} (a) Free energies of each macrostate relative to the hybridized state that is assigned a free energy of zero. (b) Discrepancy between the macrostate free energy predictions of the MSM relative to the NN model. With the exception of the F4 frayed state, the NN model predicts lower free energies than the MSM predictions.}
        \label{fig:NN_table}
	\end{center}
\end{figure}

@@

The GC-core MSM represents a departure from the relatively rich and complex kinetic networks dominated by out-of-register shifted states to a much simpler one dominated by fraying (Fig.~\ref{fig:allseq_table}c). The MSM contains only three states -- hybridized H, dehybridized D, and frayed F4. The F4 state is unique to GC-core and contains up to six WC pairs -- the two central G:C core pairs and as many as four A:T pairs on one side or other of the core, while the other run of four A:T pairs remains free. (As expected by symmetry, the particular AT run that is made vs.\ free occurs with equal probability.) Although partially frayed states containing less than four free A:T bases on either end of the duplex are common, these tend to inter-convert faster than the lag time and are not metastable within our MSM. Our model reveals the absence of any direct hybridization or dehybridization transitions between the H and D states, with all pathways passing through the frayed state F4. Previous studies would suggest that hybridization of this sequence should proceed via a zippering mechanism, wherein upon formation of the strong central G:C WC base pairings the duplex helix should rapidly assemble in a middle-out fashion.~\citep{Romano2013DNADependence, Yin2011KineticsHybridization} Our results are partially consistent with this expectation, but reveal the frayed state F4 to be unexpectedly metastable, serving as a long-lived state with an mean life time of 1.8 ns. The stability of the state is attributable to the enthalpic stabilization offered by the up to six WC pairs and the entropic stabilization associated with the configurational entropy of the two free AT-tails.

@@

%We present in Fig.~\ref{fig:T-jump_stretching}b-d our T-jump IR t-HDVE difference spectra and corresponding normalized time traces at 1600 cm$^{-1}$ and 1660 cm$^{-1}$. The signal at 1600 cm$^{-1}$ corresponds to changes in A and T ring vibrations while the signal at 1660 cm$^{-1}$ contains contributions from G and T carbonyl vibrations. Each time trace was fitted to the sum of a stretched exponential and two exponentials $S(t) = A\exp{(-(t/\tau^{\mathrm{fast}})^{\beta})} + B\exp{(-t/\tau_{\mathrm{slow}})}+C\exp{(-t/\tau_{\mathrm{cool}})}$. The stretched exponential describes the relaxation process from 5 ns to 1 $\mu$s, and the two exponentials describe the signal increase from 1-320 $\mu$s and signal decay from re-hybridization induced by thermal relaxation back to the initial temperature. The fitting parameters $A$, $B$, and $C$ correspond to the relative amplitudes of the three kinetic responses and the stretch factor $\beta$ corresponds to. A testable prediction of our hypothesis is that the long-lived out-of-register shifted states in AT-all should result in a significantly smaller value for the fitted $\beta$ parameter relative to those for GC-end, GC-core, and GC-mix. This hypothesis was supported by the experimental time series at both 1600 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.3 compared to $\beta_\mathrm{GC-end}$ = 0.6, $\beta_\mathrm{GC-core}$ = 0.7, and $\beta_\mathrm{GC-mix}$ = 0.6, and 1660 cm$^{-1}$, where $\beta_\mathrm{AT-all}$ = 0.4 compared to $\beta_\mathrm{GC-end}$ = 0.7, $\beta_\mathrm{GC-core}$ = 0.6, and $\beta_\mathrm{GC-mix}$ = 0.6. This result provides an experimental validation of the kinetic predictions of the MSM models.

@@

This material is based upon work supported by the National Science Foundation under Grant No.\ CHE-1841805. A.T. thanks the National Institute of General Medical Sciences of the National Institutes of Health (Award No. R01-GM118774) for support of this research. B.A. acknowledges support from the NSF Graduate Research Fellowship Program. This work was completed in part with resources provided by the University of Chicago Research Computing Center. We gratefully acknowledge computing time on the University of Chicago high-performance GPU-based cyberinfrastructure supported by the National Science Foundation under Grant No.\ DMR-1828629.
